---
title: "Data Wrangling BIOE 276 Assignment 1"
author: "Ward, A"
format: 
  pdf: default
editor: visual
---

# Activity 6: Data wrangling 1: rows and columns

## 1.1 Introducing tidyverse and the pipeline

Begin by loading packages

```{r}
## Load packages from library
library("tidyverse")
library(here)
```

Starting with a vector and stringing multiple functions with the pipeline:

```{r}
#| output: true

#Create vector
fruits <- c("apple", "apple", "orange", "orange", "banana")

#Find unique values
unique(fruits)
```

If I am wanting to apply the length function to this vector, I can either:

```{r}
#| output: true

# 1) Nest the unique function within the length function
length(unique(fruits))
```

```{r}
#| output: true

# 2) Create intermediate data objects
unique.fruits <- unique(fruits)
length(unique.fruits)
```

OR, I can use the pipeline here:

```{r}
#| output: true

#This :
fruits %>% unique()

#is equivalent to this:
unique(fruits)
```

And you can pipe together multiple:

```{r}
#| output: true

# Piping!
fruits %>% unique() %>% length()

#Can be spread across multiple lines
fruits %>% 
  unique() %>% 
  length()

```

## 1.2 Import data, apply the pipeline

Import data:

```{r}
# Import the data
cereal <- read_csv(here("cereal.csv"))

# Take a look at the first six rows
head(cereal)
```

We can use the pipeline to extract many things at once:

```{r}
#| output: true

# Start with the cereal dataframe
cereal %>%
  # This line extracts the column names from the dataframe input and creates a vector
  colnames() %>% 
  # This line calculates the "length" (the number) of those names from the vector input
  length()
```
